Proportional multi-state multiple-cohort life table model

### 2018-04-16 MSLT/MSLTdoc.Rmd
Plots compare finish
Improve labels and add graphs names

We start by cleaning the global environment (#1) to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoid the use of scientific notation (#2) and lastly we load the functions (#3). The code chunks are shown in the rmarkdown output.

```{r, echo=TRUE}
#1
rm (list = ls())
#2
options(scipen=999)
#3
source("code/functions.R")
```

@@

Initial case studies for the ITHIMR are: London, Sao Pablo, Delhi, Accra, Los Angeles and Edinburgh. Here, we will start with **greater London** given the availability of diseases epidemiology data from the GBD IHME study. For the rest of the case study cities data is available at the country level, hence, a scaling method is needed.

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.*](structure\Figure3.png){width=50%,height=30%,fig.pos="H"}

@@

Once the selections described in Table 2 are made, the option *Download CVS** in the GHDx website is selected. A prompt comes up asking for an email address. The data is sent to the designated email address (within minutes) in ZIP format, unzip and use the code below to read the data. Here, we selected data for Greater London and England. The aim is to compare and derive scaling factors as for most cities the data is not available from the GBD and country level data may be used and scaled to the city level. Note that the data input requirement for the PMSLT, except population numbers, is in rates. Therefore the scaling is to better reflect the burden of an area, this is a different issue than working with numbers (e.g. total mortality numbers, total YLDs numbers) as in the ITHIM approach.

```{r, echo=TRUE}
options(scipen=999)
GBDdata <- read.csv("data/IHME-GBD_2016_DATA-8fca12f6-1.csv", stringsAsFactors = F)
```

The following code serves to sort out the GBD data to the inputs required for the life table, disease life table and Dismod II

@@

```{r, echo=TRUE}
#Data preparation (generate age groups, change capitals to lower case)

#USE demography package to interpolate moratlity rates.

##All upper cases to lower cases

GBDdata <- mutate_all(GBDdata, funs(tolower))

##Create age categories index in GBDdata (rounding mid age), total of 20 age groups. These are the age cohorts to simulate.

GBDdata$age_cat [GBDdata$age =="under 5"] <- 2
GBDdata$age_cat [GBDdata$age =="5 to 9"] <- 7
GBDdata$age_cat [GBDdata$age =="10 to 14"] <- 12
GBDdata$age_cat [GBDdata$age =="15 to 19"] <- 17
GBDdata$age_cat [GBDdata$age =="20 to 24"] <- 22
GBDdata$age_cat [GBDdata$age =="25 to 29"] <- 27
GBDdata$age_cat [GBDdata$age =="30 to 34"] <- 32
GBDdata$age_cat [GBDdata$age =="35 to 39"] <- 37
GBDdata$age_cat [GBDdata$age =="40 to 44"] <- 42
GBDdata$age_cat [GBDdata$age =="45 to 49"] <- 47
GBDdata$age_cat [GBDdata$age =="50 to 54"] <- 52
GBDdata$age_cat [GBDdata$age =="55 to 59"] <- 57
GBDdata$age_cat [GBDdata$age =="60 to 64"] <- 62
GBDdata$age_cat [GBDdata$age =="65 to 69"] <- 67
GBDdata$age_cat [GBDdata$age =="70 to 74"] <- 72
GBDdata$age_cat [GBDdata$age =="75 to 79"] <- 77
GBDdata$age_cat [GBDdata$age =="80 to 84"] <- 82
GBDdata$age_cat [GBDdata$age =="85 to 89"] <- 87
GBDdata$age_cat [GBDdata$age =="90 to 94"] <- 92
GBDdata$age_cat [GBDdata$age =="95 plus"] <- 97

##Create age and sex categories to obtain population numbers. Population numbers from GBD are used in 
##Dismod II. For the Life table (Figure 3), the numbers may be from the synthetic population.

GBDdata$sex_age_cat <- paste(GBDdata$age_cat,GBDdata$sex, sep = "_"  )

##Convert variable "val" to numeric as its latter used in calculations.

GBDdata$val <- as.numeric(as.character(GBDdata$val))

##Generate population numbers for England and Greater London in a new data frame ("GBD_population") and then separate in dataframes for England and Greater London (this avoids repeating the same code).

GBD_population <- filter(GBDdata, measure == "deaths", cause == "all causes", metric == "rate" | metric == "number" ) %>% select(metric, age_cat, val, sex_age_cat, location)

###Generate population numbers from given number of cases and rates per 100,000 people. 
 for (i in 1:nrow(GBD_population)) {
    if (GBD_population$metric[i] == "number") {
      GBD_population$val_pop[i] <- GBD_population$val[i] * 100000/ GBD_population$val[i + 2]}
   else {GBD_population$val_pop[i] <- NA}
 }
##Remove rows with zero
GBD_population <- GBD_population[!is.na(GBD_population$val_pop),]

##Keep relevant variables

GBD_population <- filter(GBD_population) %>% select(sex_age_cat, val_pop, location)

##Create data frames for Greater London and England population to be later used as inputs for PMSLT calculations

GBD_population_GL <-  filter(GBD_population, location == "greater london") %>% select(sex_age_cat, val_pop, location)

GBD_population_England <-  filter(GBD_population, location == "england") %>% select(sex_age_cat, val_pop, location)

##Check population total numbers

GreaterLondon <- sum(GBD_population_GL$val_pop)
England <- sum(GBD_population_England$val_pop)

##Generate data frames for England and Grater London with per person rates (per 100,000 in original data).

#England
GBDEngland <- filter(GBDdata, location == "england" & metric == "rate") %>% select(measure, location, sex, age, metric, cause, val, age_cat)

GBDEngland$one_rate <- GBDEngland$val/100000

#Greater London (GL)
GBDGL <- filter(GBDdata, location == "greater london" & metric == "rate") %>% select(measure, location, sex, age, metric, cause, val, age_cat)
GBDGL$one_rate <- GBDGL$val/100000
```

These graphs are to compare GBD data for England and Greater London.

@@

for (sex in p_sex){
  for (cause in p_cause) {
    for (measure in p_measure) {
      p1 <- plot_GBD(in_data1 = GBDEngland, in_data2 = GBDGL, in_sex = sex, in_cause = cause, in_measure = measure)
      dev.new()
      print(p1)

    }
  }
}

##This works, check what is the issue with the loop. Seems that the problem was with colon and rectum cancer mispelling.

p_test <- plot_GBD(in_data1 = GBDEngland, in_data2 = GBDGL, in_sex = "male", in_cause = "colon and rectum cancer", in_measure = "deaths")
```

>>>>

We start by cleaning the global environment (1) to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoid the use of scientific notation (2) and lastly we load the functions (3). The code chunks are shown in the rmarkdown output.

1) Clearn Global Environment
```{r, echo=TRUE}
rm (list = ls())
```
2) Avoid scientific notation
```{r, echo=TRUE}
options(scipen=999)
```
3) Load functions 
```{r, echo=TRUE}
source("code/functions.R")
```

@@

Initial case studies for the ITHIMR are: London, Sao Pablo, Delhi, Accra, Los Angeles and Edinburgh. Here, we will start with **Greater London** given the availability of disease epidemiology data from the GBD IHME study. For the rest of the case study cities data is available at the country level, hence, a scaling method is needed to reflect the local burden of disease.

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario.*](structure\Figure3.png){width=50%,height=30%,fig.pos="H"}

@@

Once the selections described in Table 2 are made, the option *Download CVS** in the GHDx website is selected. A prompt comes up asking for an email address. The data is sent to the designated email address (within minutes) in ZIP format, unzip and use the code below to read the data (4). Here, we selected data for Greater London and England. The aim is to compare and derive scaling factors as for most cities the data is not available from the GBD and country level data may be used and scaled to the city level. Note that the data input requirement for the PMSLT, except population numbers, is in rates. Therefore the scaling is to better reflect the burden of an area, this is a different issue than working with numbers (e.g. total mortality numbers, total YLDs numbers) as in the ITHIM approach.

4) Read GBD data
```{r, echo=TRUE}
GBDdata <- read.csv("data/IHME-GBD_2016_DATA-8fca12f6-1.csv", stringsAsFactors = F)
```

The following codes serves to sort out the GBD data to the inputs required for the life table, disease life table and Dismod II

@@

5) Change all uper cases to lower case
```{r, echo=TRUE}
GBDdata <- mutate_all(GBDdata, funs(tolower))
```

6) Create age categories index in GBDdata (rounding mid age), total of 20 age groups. These are the age cohorts to simulate. 
```{r, echo=TRUE}
GBDdata$age_cat [GBDdata$age =="under 5"] <- 2
GBDdata$age_cat [GBDdata$age =="5 to 9"] <- 7
GBDdata$age_cat [GBDdata$age =="10 to 14"] <- 12
GBDdata$age_cat [GBDdata$age =="15 to 19"] <- 17
GBDdata$age_cat [GBDdata$age =="20 to 24"] <- 22
GBDdata$age_cat [GBDdata$age =="25 to 29"] <- 27
GBDdata$age_cat [GBDdata$age =="30 to 34"] <- 32
GBDdata$age_cat [GBDdata$age =="35 to 39"] <- 37
GBDdata$age_cat [GBDdata$age =="40 to 44"] <- 42
GBDdata$age_cat [GBDdata$age =="45 to 49"] <- 47
GBDdata$age_cat [GBDdata$age =="50 to 54"] <- 52
GBDdata$age_cat [GBDdata$age =="55 to 59"] <- 57
GBDdata$age_cat [GBDdata$age =="60 to 64"] <- 62
GBDdata$age_cat [GBDdata$age =="65 to 69"] <- 67
GBDdata$age_cat [GBDdata$age =="70 to 74"] <- 72
GBDdata$age_cat [GBDdata$age =="75 to 79"] <- 77
GBDdata$age_cat [GBDdata$age =="80 to 84"] <- 82
GBDdata$age_cat [GBDdata$age =="85 to 89"] <- 87
GBDdata$age_cat [GBDdata$age =="90 to 94"] <- 92
GBDdata$age_cat [GBDdata$age =="95 plus"] <- 97
```

7) Create age and sex categories to obtain population numbers. Population numbers from GBD are used in Dismod II. For the Life table (Figure 3), the numbers may be from the synthetic population. 
```{r, echo=TRUE}
GBDdata$sex_age_cat <- paste(GBDdata$age_cat,GBDdata$sex, sep = "_"  )
```
8) Conver string variables to numberic to do calculations.
```{r, echo=TRUE}
GBDdata$val <- as.numeric(as.character(GBDdata$val))
```
9) Generate population numbers for England and Greater London in a new data frame ("GBD_population") and then separate in dataframes for England and Greater London (this avoids repeating the same code). Population numbers are derived from rates per 100,000 and total numbers of cases. 
```{r, echo=TRUE}
GBD_population <- filter(GBDdata, measure == "deaths", cause == "all causes", metric == "rate" | metric == "number" ) %>% select(metric, age_cat, val, sex_age_cat, location)
```
10) Generate population numbers from given number of cases and rates per 100,000 people. 
```{r, echo=TRUE}
 for (i in 1:nrow(GBD_population)) {
    if (GBD_population$metric[i] == "number") {
      GBD_population$val_pop[i] <- GBD_population$val[i] * 100000/ GBD_population$val[i + 2]}
   else {GBD_population$val_pop[i] <- NA}
 }
```
11) Remove rows with zero
```{r, echo=TRUE}
GBD_population <- GBD_population[!is.na(GBD_population$val_pop),]
```
12) Keep relevant variables
```{r, echo=TRUE}
GBD_population <- filter(GBD_population) %>% select(sex_age_cat, val_pop, location)
```
13) Create data frames for Greater London and England population to be later used as inputs for PMSLT calculations
```{r, echo=TRUE}
GBD_population_GL <-  filter(GBD_population, location == "greater london") %>% select(sex_age_cat, val_pop, location)

GBD_population_England <-  filter(GBD_population, location == "england") %>% select(sex_age_cat, val_pop, location)
```
12) Check population total numbers
```{r, echo=TRUE}
GreaterLondon <- sum(GBD_population_GL$val_pop)
England <- sum(GBD_population_England$val_pop)
```

13) Generate data frames for England and Grater London with per person rates (per 100,000 in original data).  
```{r, echo=TRUE}
GBDEngland <- filter(GBDdata, location == "england" & metric == "rate") %>% select(measure, location, sex, age, metric, cause, val, age_cat)
GBDEngland$one_rate <- GBDEngland$val/100000
GBDGL <- filter(GBDdata, location == "greater london" & metric == "rate") %>% select(measure, location, sex, age, metric, cause, val, age_cat)
GBDGL$one_rate <- GBDGL$val/100000
```

14) Generate graphs are to compare GBD data for England and Greater London.

@@

for (sex in p_sex){
  for (cause in p_cause) {
    for (measure in p_measure){

# Exclude breast_cancer for Males
      if (sex == "male" && cause == "breast cancer" || cause == "all causes" && measure == "prevalence" || cause == "all causes" && measure == "incidence"){
        p1 <- NULL

      }
      else {
      p1 <- plot_GBD(in_data1 = GBDEngland, in_data2 = GBDGL, in_sex = sex, in_cause = cause, in_measure = measure)

      }
    }
  }
}
